---
title: "confitns"
author: "Dominik Kavrik"
date: "2023-03-09"
output: html_document
---
```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
theme_set(theme_light())
```

# Simulate a population:

```{r}
N = 100000
x = rnorm(N, 1 , 2)
y = 15 + 30*x + rnorm(N, 2, 60)
population <- tibble(y,x) 
```

# pick a random sample and visualize it

```{r}
n <- 500
sample <- sample_n(population, n)
```

```{r}
sample %>% 
  ggplot(aes(x,y)) +
  geom_point() + 
  labs(title = "Sample Scatterplot",
       subtitle = paste(n, "observations"))
```
# add another layer with regression line:


```{r}
sample %>% 
  ggplot(aes(x,y)) +
  geom_point() + 
  labs(title = "Sample Scatterplot",
       subtitle = paste(n, "observations")) +
  geom_smooth(method = "lm")
```

Notice the grey area. That is the simple regression confidence interval.
CI is a function of x and is non-linear.



The line is the fitted value defined as $$\hat{y_i} = \hat{\beta_0} + \hat{\beta_1}x_i$$

We can actually calculate the CI from scratch using the following formula:

$$CI = \hat{y} \pm t_{\alpha/2 \; n-k-1} \times \sqrt{  \hat{\sigma}^2 \bigg( \frac{1}{n} + \frac{(x_i - \bar{x})^2}{\sum_{i=1}^n(x_i - \bar{x})^2}\bigg)}$$

first we need to estimate the model: 

$$y_i = \beta_0 + \beta_1x_i + u_i$$

```{r}
model <- lm(y ~ x, data = sample)
summary(model)
```

$$y_i = \hat{\beta_0}+\hat{\beta_1}x_i+\hat{u_i}$$
$$y_i = 15.79 + 30.55 x_i + \hat{u_i}$$
$$\hat{y}_i = \hat{\beta_0}+\hat{\beta_1}x_i$$

$$\hat{u}_i = y_i - \hat{y_i}$$



and we need to get the estimated variance of the estimator:

$$\hat{\sigma}^2 = \frac{1}{n-2} \sum_{i=1}^n(\hat{u}_i^2)$$

```{r}
sigma_sq_hat = 1/(n-2)*sum(model$residuals^2)
```


We can easily coerce the model residuals into dataframe using `{broom}` package

```{r}
CI_data <- broom::augment(model) %>% 
  select(y,x,.fitted) %>% 
  rename(fitted = ".fitted") %>% 
  mutate(upper = fitted + 1.96 * sqrt(sigma_sq_hat* (1/n + (x - mean(x))^2 / ( sum((x-mean(x))^2)) ) )) %>% 
  mutate(lower = fitted - 1.96 * sqrt(sigma_sq_hat* (1/n + (x - mean(x))^2 / ( sum((x-mean(x))^2)) ) ))
```



```{r}
head(CI_data)
```

```{r}
CI_data %>%
  pivot_longer(names_to = "names", values_to = "values", 3:5) %>% 
  ggplot(aes(x = x, y=values, color = names, linetype=names))+
  geom_line()+
  scale_linetype_manual(values=c("solid", "dashed", "dashed"))
```
Now see that it really is the outline of the CI given by the `geom_smooth()`

```{r}
CI_data %>%
  pivot_longer(names_to = "names", values_to = "values", 3:5) %>% 
  ggplot(aes(x = x, y=values, color = names, linetype=names))+
  geom_line()+
  scale_linetype_manual(values=c("solid", "dashed", "dashed"))+
  geom_point(aes(x,y), color = "black", alpha = 0.5, shape = "o")+
  geom_smooth(aes(x,y), method = "lm")
```

